A DPG method for compressible flow problems

### 2012-06-26 docs/Jesse/Proposal/proposalIntro.tex
adding updates to proposal tex files

Over the last three decades, Computational Fluid Dynamics (CFD) simulations have become commonplace as a tool in the engineering and design of high-speed aircraft.  Wind tunnel experiments are often complemented by computational simulations, and CFD technologies have proved very useful in both the reduction of aircraft development cycles, and in the simulation of conditions difficult to reproduce experimentally.  Great advances have been made in the field since its introduction, especially in areas of meshing, computer architecture, and solution strategies.  Despite this, there still exist many computational limitations in existing CFD methods:  
\begin{itemize}
\item \textbf{Higher order methods :} Higher order methods stand to offer large computational savings through a more efficient use of discrete degrees of freedom.  However, there are very few working higher-order CFD codes in existence, and most higher order methods tend to degrade to first-order accuracy near shocks. The use of higher order codes to solve the steady state equations is even rarer, where convergence of discrete nonlinear steady equations is a tricky issue \cite{BoeingHigherOrder}.  
\item \textbf{Automatic adaptivity :} The use of adaptive meshes is crucial to many CFD applications, where the solution often exhibits very localized sharp gradients and shocks.  Good resolution for such problems under uniform meshes is computationally prohibitive and impractical for most physical regimes of interest.  However, the construction of ``good" meshes is a difficult task, usually requiring a-priori knowledge of the form of the solution \cite{andersonCFDBook}.  An alternative to such is the construction of automatically adaptive schemes; such methods usually begin with a coarse mesh and refine based on the minimization of some error.  However, this task is difficult, as the convergence of numerical methods for problems in CFD is notoriously sensitive to mesh quality.  Additionally, the use of adaptivity becomes even more difficult in the context of higher order and $hp$ methods \cite{BoeingHigherOrder}.  
\end{itemize}
Both of these issues are tied to the notion of \emph{robustness}.  We define robustness loosely as the degradation of the quality of numerical solutions with respect to a given problem parameter.  In the context of CFD, the parameter of interest is the Reynolds number (the nondimensional equivalent of viscosity) --- for typical physical conditions of interest for the compressible Navier-Stokes equations, the Reynolds number is extremely high, on the order of 1e7, yielding solutions with two vastly different scales - inviscid phenomena at an $O(1)$ scale, and $O(1e-7)$ viscous phenomena.

@@

Standard numerical methods tend to perform poorly across the board for the class of PDEs known as singular perturbation problems. These problems are often characterized by a parameter that may be either very small or very large in the context of physical problems.  An additional complication of singular perturbation problems is that very often, in the limiting case of the parameter blowing up or decreasing to zero, the PDE itself will change types (e.g.\ from elliptic to hyperbolic).  A canonical example of a singularly perturbed problem is the convection-diffusion equation on domain $\Omega \in \mathbb{R}^3$,
\[
\div \left(\beta u\right) - \epsilon \Delta u = f.
\]
The equation represents the change in the scalar quantity $u$, representing the concentration of a quantity in a given medium, taking into account both convective and diffusive effects. $\beta \in \mathbb{R}^3$ specifies the direction and magnitude of convection, while the singular perturbation parameter $\epsilon$ represents the diffusivity of the medium. In the limit of an inviscid medium as $\epsilon\rightarrow 0$, the equation changes types, from elliptic to hyperbolic, and from second order to first order.

@@

\begin{figure}[!h]
\centering
\includegraphics[scale=.4]{figs/GalerkinOscTight.png}
\caption{Oscillations in the 1D finite element solution of the convection-diffusion equation for small diffusion \cite{GalerkinOsc}. Standard finite volume and finite difference methods exhibit similar behavior.}
\label{fig:GalerkinOsc}
\end{figure}
We consider for now the Galerkin finite element method as applied to convection-dominated diffusion.  Standard finite element methods (as well as standard finite volume and finite difference methods) perform poorly for the case of small $\epsilon$.  The poor performance of the finite element method for this problem is reflected in the bound on the error in the finite element solution --- under the standard Bubnov-Galerkin method with $u\in H^1(0,1)$, we have the bound given in \cite{roos2008robust}:
\[
\|u-u_h\|_\epsilon \leq C \inf_{w_h}\|u-w_h\|_{H^1(0,1)},
\]
for $\|u\|_\epsilon^2 \coloneqq \|u\|_{L^2}^2 + \epsilon \|u'\|_{L^2}^2$, with $C$ independent of $\epsilon$. An alternative formulation of the above bound is 
\[
\|u-u_h\|_{H^1(0,1)} \leq C(\epsilon) \inf_{w_h}\|u-w_h\|_{H^1(0,1)},
\]
where $C(\epsilon)$ grows as $\epsilon\rightarrow 0$. The dependence of the constant $C$ on $\epsilon$ is what we refer to as a \textit{loss of robustness} --- as the singular perturbation parameter $\epsilon$ decreases, our finite element error is bounded more and more loosely by the best approximation error.  As a consequence, the finite element solution can diverge significantly from the best finite element approximation of the solution for very small values of $\epsilon$.  For exam ple, Figure~\ref{fig:GalerkinOsc} shows an example of how, on a coarse mesh, and for small values of $\epsilon$, the Galerkin approximation of the solution to the convection-diffusion equation with a boundary layer develops spurious oscillations everywhere in the domain, even where the best approximation error is small.  These oscillations grow in magnitude as $\epsilon \rightarrow 0$, eventually polluting the entire solution.

@@

From the perspective of the full Navier-Stokes equations, this loss of robustness is doubly problematic.  Not only will any nonlinear solution suffer from similar non-physical oscillations, but nonlinear solvers themselves may fail to yield a solution due to such instabilities --- a nonlinear solution is almost always computed by solving a series of linear problems whose solutions will converge to the nonlinear solution under appropriate assumptions, and the presence of such oscillations in each linear problem can cause the solution convergence to slow significantly or even diverge.

Our goal is to develop a \emph{stable, higher order adaptive scheme} for the steady compressible laminar Navier-Stokes equations in transonic/supersonic regimes that is \emph{robust for very small viscosity}.  In particular, we hope to present a method for which $hp$-adaptivity can be applied to problems in compressible flow.

@@

For the past half-century, problems in CFD have been solved using a multitude of methods, many of which are physically motivated, and thus applicable only to a small number of problems and geometries. We consider more general methods, whose framework is applicable to different problems in CFD; however, our specific focus will be on the nonlinear shock problems of compressible aerodynamics. Broadly speaking, the most popular general methods include (in historical order) finite difference methods, finite volume methods, and finite element methods.

@@

For linear problems, finite difference (FD) methods approximate derivatives based on interpolation of pointwise values of a function.  FD methods were popularized first by Lax, who introduced the concepts of the monotone scheme and numerical flux. For the conservation laws governing compressible aerodynamics, FD methods approximated the conservation law, using some numerical flux to reconstruct approximations to the derivative at a point. Finite volume (FV) methods are similar to finite difference methods, but approximate the integral version of a conservation law as opposed to the differential form. FD and FV have roughly the same computational cost/complexity; however, the advantage of FV methods over FD is that FV methods can be used on a much larger class of discretizations than FD methods, which require uniform or smooth meshes.

For nonlinear shock problems, the solution often exhibits sharp gradients or discontinuities, around which the solution would develop spurious Gibbs-type oscillations. Several ideas were introduced to deal with oscillations in the solution near a sharp gradient or shock: artificial diffusion, total variation diminishing (TVD) schemes, and slope limiters. However, each method had its drawback, either in terms of loss of accuracy, dimensional limitations, or problem-specific parameters to be tuned \cite{Shu:Lectures}. Harten, Enquist, Osher and Chakravarthy introduced the essentially non-oscillatory (ENO) scheme in 1987 \cite{ENO}, which was improved upon with the weighted essentially non-oscillatory (WENO) scheme in \cite{WENO}. WENO remains a popular choice today for both finite volume and finite difference schemes. Most of these methods can be interpreted as adding some specific artificial diffusion to the given numerical scheme.  We refer to such schemes as \emph{modified equation} methods, as the exact solution no longer satisfies the discrete system due to the presence of additional artificial diffusion.

@@

Historically, the Galerkin method has been very successfully applied to a broad range of problems in solid mechanics, for which the variational problems resulting from the PDE are symmetric and coercive (positive-definite). It is well known that the finite element method produces optimal or near-optimal results for such problems, with the finite element solution matching or coming close to the best approximation of the solution in the finite element space. However, standard Galerkin methods tend to perform poorly for singular perturbation problems, developing instabilities when the singular perturbation parameter is very small.

@@

All Galerkin methods involve both trial (approximating) and test (weighting) functions.   Standard Galerkin methods, where these trial and test functions are taken from the same space, are referred to as Bubnov-Galerkin methods.  A Petrov-Galerkin method refers to any Galerkin method other than a classical Bubnov-Galerkin method.  The Streamline Upwind Petrov Galerkin (SUPG) method is a stabilization method for $H^1$-conforming finite elements, the idea of which was originally motivated by artificial diffusion techniques in finite differences.  In particular, for the homogeneous 1D convection-diffusion equation, it is possible to recover, under a finite difference method, the exact solution at nodal points by adding an ``exact" artificial diffusion based on the mesh size $h$ and the magnitudes of the convection $\beta$ and the viscosity $\epsilon$.

The idea of ``exact" artificial viscosity was adapted to finite elements not through the direct modification of the equations, but through the test functions and weighting of the residual.  In particular, finite element and Galerkin methods are often referred to as ``weighted residual" methods, since the starting point of both is to multiply the residual by a particular test, or weighting, function.  Standard Bubnov-Galerkin methods simply choose these weighting functions to be the same as the the basis functions used to approximate the solution.

We will introduce the SUPG method at an abstract level for illustrative purposes only.  Further details and perspectives on the SUPG method can be found in \cite{SUPG}, as well as in an upcoming book by Hughes.  The convection-diffusion equation can be written as follows:
\[
Lu = \left(L_{\rm adv} + L_{\rm diff} \right) u = f,
\]
where $L_{\rm adv}u \coloneqq \div \left(\beta u\right)$ is the first order advective operator, and $L_{\rm diff}u \coloneqq \epsilon \Delta u$ is the second-order diffusive operator (then, roughly speaking, $L_{\rm diff} u = 0$).  Let us assume $u$ to be a linear combination of piecewise-linear basis functions $\phi_i, i = 0,\ldots,N$.  The SUPG method is then to solve
\[
\int_\Omega L u v + \int_\Omega \tau \left(L_{\rm adv}v\right)\left( L u - f\right) = \int_\Omega f v 
\]
for where $\tau$ is the SUPG parameter.  For uniform meshes in 1D, $\tau$ is chosen such that, for $f=0$, the matrix system resulting from SUPG is exactly equal to the finite difference system under ``exact" artificial diffusion.  However, unlike the artificial diffusion method, for $f\neq 0$, the SUPG method still delivers optimal stabilization.  In fact, the SUPG finite element solution in 1D is nothing less than the nodal interpolant and the best $H^1$ approximation of the exact solution, as seen in Figure~\ref{fig:SUPG}.

@@

This addition of residual-based stabilization terms can be interpreted as a modification of the test functions as well.  For SUPG, the formulation can equivalently be written as
\[
\int_\Omega\left( Lu-f\right) v_i = \int_\Omega f v_i, \quad i = 0,\ldots,N
\]
where $v_i$ is defined as
\[
v_i = \phi_i(x) + \tau L_{\rm adv} \phi_i.  
\]
In other words, the test function $v_i$ is a perturbation of the basis function $\phi_i$ by a scaled advective operator applied to $\phi_i$.  For a linear $C^0$ basis function (the ``hat" function), this naturally leads to a bias in the upwind or streamline direction of the flow $\beta$, as seen in Figure~\ref{fig:SUPG}.

@@

Discontinuous Galerkin (DG) methods form a subclass of FEM; first introduced by Reed and Hill in \cite{Reed:73}, these methods were later analyzed by Cockburn and Shu \cite{CockburnShu:DG} and have rapidly gained popularity for CFD problems. Advantages of DG methods include the local conservation property, easily modified local orders of approximation, easy adaptivity in both $h$ and $p$, and efficient parallelizability. Rather than having a continuous basis where the basis function support spans several element cells, DG opts instead for a discontinuous, piecewise polynomial basis, where, like FV schemes, a \emph{numerical flux} facilitates communication between neighboring elements (unlike FV methods, however, there is no need for a reconstruction step).

The formal definition of the numerical flux (attributed to Peter Lax) on an element boundary is some function of the values on the edges of both the neighboring elements.  An additional reason for the popularity of DG methods is that they can be interpreted as stabilized FE methods through appropriate choices of this numerical flux \cite{Brezzi20063293}. We will illustrate this with the steady convection equation in 1D ---
\begin{align*}
\pd{\left(\beta(x) u\right)}{x} = f, \quad u(0) &= u_0.
\end{align*}
The DG formulation is derived by multiplying by a test function $v$ with support only on a single element $K = [x_K,x_{K+1}]$ and integrating by parts. The boundary term is left alone, such that the local formulation is 
\[
\left.\beta u v\right|_{x_K}^{x_{K+1}} + \int_K -\beta u \pd{v}{x} = \int_K f v,
\]
and the global formulation is recovered by summing up all element-wise local formulations. However, the boundary term in the local formulation is presently ill-defined, as both $u$ and $v$ are dual-valued over element boundaries. Consequently, we make the choice to define the values of $u$ on the boundary (the \emph{traces} of $u$) as
\begin{align*}
u(x_K) \coloneqq u(x_K^-), \quad u(x_{K+1}) \coloneqq u(x_{K+1}^-),
\end{align*}
where $u(x_K^-)$ is the value of $u$ at $x_K$ as seen from the left, and $u(x_K^+)$ the value as seen from the right. Similarly, the traces of $v$ are defined to be
\begin{align*}
v(x_K) \coloneqq v(x_K^+), \quad v(x_{K+1}) \coloneqq v(x_{K+1}^-),
\end{align*}
We will refer to $v(x_K^+)$ as the upwind value of $v(x_K)$, and refer to DG under this specific choice of traces as upwind DG.  This specific choice of $v(x_K)$ as the upwind value is crucial; similarly to SUPG, the upwind DG emphasizes the test function in the direction of convection and changes the way the residual is measured.  As it turns out, the performance of DG for convection-type problems is closely tied to this upwinding --- choosing the value of $v(x_K)$ to be the downwind value $v(x_K^-)$ leads to an unstable method, while choosing $v(x_K)$ to be the average of the upwind and downwind values leads to a DG method with suboptimal stability properties, similar to an $H^1$-conforming continuous Galerkin approximation\cite{Brezzi20063293}.\footnote{For second-order convection-diffusion problems with small diffusion, the additional regularity imparted by choice of the DG numerical flux is often insufficient, and SUPG-type stabilization is also applied.}

@@

HDG methods are an active topic of current research, since they address several criticisms of common DG methods (large number of degrees of freedom, globally coupled degrees of freedom, complicated/inefficient implementation procedures, suboptimal convergence of approximate fluxes). Note that HDG methods still fall under the category of stabilized methods --- stabilization techniques are employed through the choice of the HDG numerical flux, which involves some stabilization parameter $\tau$.

@@

This PhD thesis proposal will proceed in four main parts:
\begin{enumerate}
\item Introduce the abstract DPG method as a minimum residual method for linear problems.
\item Formulate a DPG method addressing the issue of robustness for the model problem of convection-dominated diffusion.
\item Apply the DPG method to singularly perturbed nonlinear problems in CFD, discussing preliminary results for the Burgers and compressible Navier-Stokes equations. 
\item Introduce completed and proposed work in concentration areas A, B, and C. 
\end{enumerate}

>>>>

Over the last three decades, Computational Fluid Dynamics (CFD) simulations have become commonplace as a tool in the engineering and design of high-speed aircraft.  Wind tunnel experiments are often complemented by computational simulations, and CFD technologies have proved very useful in both the reduction of aircraft development cycles and the simulation of experimentally difficult conditions.  Great advances have been made in the field since its introduction, especially in areas of meshing, computer architecture, and solution strategies.  Despite this, there still exist many computational limitations in existing CFD methods:  
\begin{itemize}
\item \textbf{Higher order methods :} Higher order methods stand to offer large computational savings through a more efficient use of discrete degrees of freedom.  However, there are very few working higher-order CFD codes in existence, and most higher order methods tend to degrade to first-order accuracy near shocks. The use of higher order codes to solve the steady state equations is even rarer, where convergence of discrete nonlinear steady equations is a tricky issue \cite{BoeingHigherOrder}.  
\item \textbf{Automatic adaptivity :} The use of adaptive meshes is crucial to many CFD applications, where the solution can exhibit very localized sharp gradients and shocks.  Good resolution for such problems under uniform meshes is computationally prohibitive and impractical for most physical regimes of interest.  However, the construction of ``good" meshes is a difficult task, usually requiring a-priori knowledge of the form of the solution \cite{andersonCFDBook}.  An alternative set of strategies are automatically adaptive schemes; such methods usually begin with a coarse mesh and refine based on the minimization of some error.  However, this task is difficult, as the convergence of numerical methods for problems in CFD is notoriously sensitive to mesh quality.  Additionally, the use of adaptivity becomes even more difficult in the context of higher order and $hp$ methods \cite{BoeingHigherOrder}.  
\end{itemize}
Both of these issues are tied to the notion of \emph{robustness}.  We define robustness loosely as the degradation of the quality of numerical solutions with respect to a given problem parameter.  In the context of CFD simulations, the parameter of interest is the Reynolds number (the nondimensional equivalent of the inverse of the viscosity) --- for typical physical conditions of interest for the compressible Navier-Stokes equations, the Reynolds number is extremely high, on the order of 1e7, yielding solutions with two vastly different scales - inviscid phenomena at an $O(1)$ scale, and $O(1e-7)$ viscous phenomena.

@@

Standard numerical methods tend to perform poorly across the board for the class of PDEs known as singular perturbation problems; these problems are often characterized by a parameter that may be either very small or very large.  An additional complication of singular perturbation problems is that very often, in the limiting case of the parameter blowing up or decreasing to zero, the PDE itself will change types (e.g.\ from elliptic to hyperbolic).  A canonical example of a singularly perturbed problem is the convection-diffusion equation in domain $\Omega \in \mathbb{R}^3$,
\[
\div \left(\beta u\right) - \epsilon \Delta u = f.
\]
The equation models the steady-state distribution of the scalar quantity $u$, representing the concentration of a quantity in a given medium, taking into account both convective and diffusive effects. Vector $\beta \in \mathbb{R}^3$ specifies the direction and magnitude of convection, while the singular perturbation parameter $\epsilon$ represents the diffusivity of the medium. In the limit of an inviscid medium as $\epsilon\rightarrow 0$, the equation changes types, from elliptic to hyperbolic, and from second order to first order.

@@

\begin{figure}[!h]
\centering
\includegraphics[scale=.4]{figs/GalerkinOscTight.png}
\caption{Oscillations in the 1D finite element solution of the convection-diffusion equation for small diffusion \cite{GalerkinOsc}. Standard finite volume and finite difference methods exhibit similar behavior.}
\label{fig:GalerkinOsc}
\end{figure}
We consider for now the Galerkin finite element method as applied to convection-dominated diffusion.  Standard finite element methods (as well as standard finite volume and finite difference methods) perform poorly for the case of small $\epsilon$.  The poor performance of the finite element method for this problem is reflected in the bound on the error in the finite element solution --- under the standard Bubnov-Galerkin method with $u\in H^1(0,1)$, we have the bound given in \cite{roos2008robust}:
\[
\|u-u_h\|_\epsilon \leq C \inf_{w_h}\|u-w_h\|_{H^1(0,1)},
\]
for $\|u\|_\epsilon^2 \coloneqq \|u\|_{L^2}^2 + \epsilon \|u'\|_{L^2}^2$, with $C$ independent of $\epsilon$. An alternative formulation of the above bound is 
\[
\|u-u_h\|_{H^1(0,1)} \leq C(\epsilon) \inf_{w_h}\|u-w_h\|_{H^1(0,1)},
\]
where $C(\epsilon)$ grows as $\epsilon\rightarrow 0$. The dependence of the constant $C$ on $\epsilon$ is what we refer to as a \textit{loss of robustness} --- as the singular perturbation parameter $\epsilon$ decreases, our finite element error is bounded more and more loosely by the best approximation error.  As a consequence, the finite element solution can diverge significantly from the best finite element approximation of the solution for very small values of $\epsilon$.  For example, Figure~\ref{fig:GalerkinOsc} shows an example of how, on a coarse mesh, and for small values of $\epsilon$, the Galerkin approximation of the solution to the convection-diffusion equation with a boundary layer develops spurious oscillations everywhere in the domain, even where the best approximation error is small.  These oscillations grow in magnitude as $\epsilon \rightarrow 0$, eventually polluting the entire solution.\footnote{For nonlinear shock problems, the solution often exhibits sharp gradients or discontinuities, around which the solution would develop spurious Gibbs-type oscillations. These are a result of underresolution of the solution, and are separate from the oscillations resulting from a lack of robustness.}

@@

From the perspective of the full Navier-Stokes equations, this loss of robustness is doubly problematic.  Not only will any nonlinear solution suffer from similar non-physical oscillations, but nonlinear solvers themselves may fail to yield a solution due to such instabilities.  A nonlinear solution is almost always computed by solving a series of linear problems whose solutions will converge to the nonlinear solution under appropriate assumptions, and the presence of such oscillations in each linear problem can cause the solution convergence to slow significantly or even diverge.

Our aim is to develop a \emph{stable, higher order adaptive scheme} for the steady compressible laminar Navier-Stokes equations in transonic/supersonic regimes that is \emph{robust for very small viscosity}.  In particular, we hope to present a method for which $hp$-adaptivity can be applied to problems in compressible flow.  The goal of this dissertation will be to develop a mathematical theory demonstrating the robustness in $\epsilon$ of our method for singularly perturbed convection-diffusion problems, and to demonstrate its feasibility as a CFD solver by applying it to several benchmark problems.  
\todo{Check with LD if it's what he's looking for}

@@

For the past half-century, problems in CFD have been solved using a multitude of methods, many of which are physically motivated, and thus applicable only to a small number of problems and geometries. We consider more general methods, whose framework is applicable to a larger set of problems; however, our specific focus will be on the problems of compressible aerodynamics involving small-scale viscous phenomena (i.e.\ boundary layers and, if present, shock waves). Broadly speaking, the most popular general methods include (in historical order) finite difference methods, finite volume methods, and finite element methods.

@@

For linear problems, finite difference (FD) methods approximate derivatives based on interpolation of pointwise values of a function.  In the context of conservation laws, FD methods were popularized first by Lax, who introduced the concepts of the monotone scheme and numerical flux. For the conservation laws governing compressible aerodynamics, FD methods approximate the conservation law, using some numerical flux to reconstruct approximations to the derivative at a point. Finite volume (FV) methods are similar to finite difference methods, but approximate the integral version of a conservation law as opposed to the differential form. FD and FV have roughly the same computational cost/complexity; however, the advantage of FV methods over FD is that FV methods can be used on a much larger class of problems and geometries than FD methods, which require uniform or smooth structured meshes.

Several ideas were introduced to deal with oscillations in the solution near a sharp gradient or shock: artificial diffusion, total variation diminishing (TVD) schemes, and slope limiters. However, each method had its drawback, either in terms of loss of accuracy, dimensional limitations, or problem-specific parameters to be tuned \cite{Shu:Lectures}. Harten, Enquist, Osher and Chakravarthy introduced the essentially non-oscillatory (ENO) scheme in 1987 \cite{ENO}, which was improved upon with the weighted essentially non-oscillatory (WENO) scheme in \cite{WENO}. WENO remains a popular choice today for both finite volume and finite difference schemes. Most of these methods can be interpreted as adding some specific artificial diffusion to the given numerical scheme.  We refer to such schemes as \emph{modified equation} methods, as the exact solution no longer satisfies the discrete system due to the presence of additional artificial diffusion terms.

@@

Historically, the Galerkin method has been very successfully applied to a broad range of problems in solid mechanics, for which the variational problems resulting from the PDE are often symmetric and coercive (positive-definite). It is well known that the finite element method produces optimal or near-optimal results for such problems, with the finite element solution matching or coming close to the best approximation of the solution in the finite element space. However, standard Galerkin methods tend to perform poorly for singular perturbation problems, developing instabilities when the singular perturbation parameter is very small.

@@

All Galerkin methods involve both trial (approximating) and test (weighting) functions.   Standard Galerkin methods, where these trial and test functions are taken from the same space, are referred to as Bubnov-Galerkin methods.  Petrov-Galerkin methods refer most often to methods where test and trial functions \emph{differ}, leading to differing test and trial spaces.\footnote{Hughes takes the more general definition of a Petrov-Galerkin method to be any Galerkin method other than a classical Bubnov-Galerkin method.}  The Streamline Upwind Petrov Galerkin (SUPG) method is a stabilization method for $H^1$-conforming finite elements, the idea of which was originally motivated by artificial diffusion techniques in finite differences.  In particular, for the homogeneous 1D convection-diffusion equation, it is possible to recover, under a finite difference method, the exact solution at nodal points by adding an ``exact" artificial diffusion based on the mesh size $h$ and the magnitudes of the convection $\beta$ and the viscosity $\epsilon$.  The idea of ``exact" artificial viscosity was adapted to finite elements not through the direct modification of the equations, but through the \emph{test} functions and weighting of the residual. \footnote{Finite element and Galerkin methods are often referred to as ``weighted residual" methods, since the starting point of both is to multiply the residual by a particular test, or weighting, function.  Standard Bubnov-Galerkin methods simply choose these weighting functions to be the same as the the basis functions used to approximate the solution.  }

We will introduce the SUPG method at an abstract level for illustrative purposes only.  Further details and perspectives on the SUPG method can be found in \cite{SUPG}, as well as in an upcoming book by Hughes.  The convection-diffusion equation can be written as follows:
\[
Lu = \left(L_{\rm adv} + L_{\rm diff} \right) u = f,
\]
where $L_{\rm adv}u \coloneqq \div \left(\beta u\right)$ is the first order advective operator, and $L_{\rm diff}u \coloneqq \epsilon \Delta u$ is the second-order diffusive operator.  Let us assume $u$ to be a linear combination of piecewise-linear basis functions $\phi_i, i = 0,\ldots,N$ (then, within each element, $L_{\rm diff} u = 0$).  The SUPG method is then to solve
\[
\int_\Omega L u v + \int_\Omega \tau \left(L_{\rm adv}v\right)\left( L u - f\right) = \int_\Omega f v 
\]
for where $\tau$ is the SUPG parameter.  For uniform meshes in 1D, $\tau$ is chosen such that, for $f=0$, the matrix system resulting from SUPG is exactly equal to the finite difference system under ``exact" artificial diffusion.  However, unlike exact artificial diffusion, for $f\neq 0$, the SUPG method still delivers optimal stabilization.  In fact, the SUPG finite element solution in 1D is nothing less than the nodal interpolant and the best $H^1_0$ approximation of the exact solution, as seen in Figure~\ref{fig:SUPG}.

@@

This addition of residual-based stabilization terms can be interpreted as a modification of the test functions as well.  For SUPG, the formulation can equivalently be written as
\[
\int_\Omega\left( Lu-f\right) v_i = \int_\Omega f v_i, \quad i = 0,\ldots,N
\]
\todo{check to make sure this is correct}
where $v_i$ is defined as
\[
v_i = \phi_i(x) + \tau L_{\rm adv} \phi_i.  
\]
In other words, the test function $v_i$ is a perturbation of the basis function $\phi_i$ by a scaled advective operator applied to $\phi_i$.  For a linear $C^0$ basis function (the ``hat" function), this naturally leads to a bias in the upwind or streamline direction of the flow $\beta$, as seen in Figure~\ref{fig:SUPG}.

@@

Discontinuous Galerkin (DG) methods form a subclass of FEM; first introduced by Reed and Hill in \cite{Reed:73}, these methods were later analyzed by Cockburn and Shu \cite{CockburnShu:DG} and have rapidly gained popularity for CFD problems. Advantages of DG methods include the local conservation property, easily modified local orders of approximation, ease of adaptivity in both $h$ and $p$, and efficient parallelizability. Rather than having a continuous basis where the basis function support spans multiple element cells, DG opts instead for a discontinuous, piecewise polynomial basis, where, like FV schemes, a \emph{numerical flux} facilitates communication between neighboring elements (unlike FV methods, however, there is no need for a reconstruction step).

The formal definition of the numerical flux (attributed to Peter Lax) on an element boundary is some function of the values on the edges of both the neighboring elements.  An additional reason for the popularity of DG methods is that they can be interpreted as stabilized FE methods through appropriate choices of this numerical flux \cite{Brezzi20063293}. We will illustrate this with the steady convection equation in 1D:
\begin{align*}
\pd{\left(\beta(x) u\right)}{x} = f, \quad u(0) &= u_0.
\end{align*}
The DG formulation is derived by multiplying by a test function $v$ with support only on a single element $K = [x_K,x_{K+1}]$ and integrating by parts. The boundary term is left alone, such that the local formulation is 
\[
\left.\beta u v\right|_{x_K}^{x_{K+1}} + \int_K -\beta u \pd{v}{x} = \int_K f v,
\]
and the global formulation is recovered by summing up all element-wise local formulations. However, the boundary term in the local formulation is presently ill-defined, as both $u$ and $v$ are dual-valued over element boundaries. Consequently, we make the choice to define the values of $u$ on the boundary (the \emph{traces} of $u$) as
\begin{align*}
u(x_K) \coloneqq u(x_K^-), \quad u(x_{K+1}) \coloneqq u(x_{K+1}^-),
\end{align*}
where $u(x_K^-)$ is the value of $u$ at $x_K$ as seen from the left, and $u(x_K^+)$ the value as seen from the right. Similarly, the traces of $v$ are defined to be
\begin{align*}
v(x_K) \coloneqq v(x_K^+), \quad v(x_{K+1}) \coloneqq v(x_{K+1}^-),
\end{align*}
For $\beta$ positive, $v(x_K^+)$ is the \emph{upwind} value of $v(x_K)$, and we refer to DG under this specific choice of traces as upwind DG.  This specific choice of $v(x_K)$ as the upwind value is crucial; similarly to SUPG, the upwind DG emphasizes the test function in the direction of convection and changes the way the residual is measured.  As it turns out, the performance of DG for convection-type problems is closely tied to this upwinding --- choosing the value of $v(x_K)$ to be the downwind value $v(x_K^-)$ leads to an unstable method, while choosing $v(x_K)$ to be the average of the upwind and downwind values leads to a DG method with suboptimal stability properties, similar to an $H^1$-conforming continuous Galerkin approximation\cite{Brezzi20063293}.\footnote{For second-order convection-diffusion problems with small diffusion, the additional regularity imparted by choice of the DG numerical flux is often insufficient, and SUPG-type stabilization is also applied.}

@@

HDG methods are an active topic of current research, since they address several criticisms of common DG methods (large number of globally coupled degrees of freedom, complicated/inefficient implementation procedures, suboptimal convergence of approximate fluxes). Note that HDG methods still fall under the category of stabilized methods --- stabilization techniques are employed through the choice of the HDG numerical flux, which involves some stabilization parameter $\tau$.

@@

This PhD thesis proposal will proceed in four main parts.  We will begin by introducing the abstract Discontinuous Petrov-Galerkin (DPG) method as a minimum residual method for linear problems and highlighting some important properties of the method.  Our next step will be to formulate and prove the robustness of a DPG method (with respect to $\epsilon$) for the model problem of convection-dominated diffusion.  Finally, we will extend and apply the DPG method to singularly perturbed nonlinear problems in CFD, presenting preliminary results for the Burgers and compressible Navier-Stokes equations.
